---
title: "1. Datos, Modelos y Software. Ejercicios"
date: "Enero 2025"
output: 
  html_document:
    toc: true
    toc_depth: 1
    theme: cerulean
---


<style>
body {
text-align: justify}
</style>

<style type="text/css">
body{/*Normal*/
     font-size: 20px;
}
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(discrtr) # A companion package for the book Introduction to Discrete Choice Analysis with `R`
library(mlogit) # Multinomial Logit Models
library(readr) # Read Rectangular Text Data
library(plotly)
```



Invoke data set Mode from package mlogit. To do this you need to first load the package. This is a data set with choices about mode of transportation. This is done as follows:

```{r}
data("Mode")
```

Once you have loaded the data set, answer the following questions:
Describe this data set. How many variables are there and of which type (i.e., categorical/quantitative)?

```{r}
glimpse(Mode)
# El conjunto de datos `Mode` contiene 22 variables.
# Incluye variables cuantitativas (como cost.car, time.bus, etc.) y categóricas (como choice).
```





How many different modes of transportation are in this data set? What is the most popular mode? What is the least popular mode?

```{r}
summary(Mode)
```
```{r}
ggplot(Mode)+
  geom_bar(aes(choice, fill=choice), color="black", show.legend = FALSE)+
# Hay 4 modos de transporte: car, carpool, bus y rail.
# El modo más popular es car.
# El menos popular es rail.
  theme_bw()
```


In general, what is the most expensive mode? The least expensive?


```{r}
ggplot(Mode)+
  geom_density(aes(cost.car, fill="car"), color="black", alpha=0.5)+
# El modo más caro en promedio es car.
# El más barato en promedio es bus.
  geom_density(aes(cost.carpool , fill="carpool"), color="black", alpha=0.5)+
  geom_density(aes(cost.bus, fill="bus"), color="black", alpha=0.5)+
  geom_density(aes(cost.rail, fill="rail"), color="black", alpha=0.5)+
  scale_fill_manual(values=c("car"= "firebrick", "carpool"="dodgerblue", "bus"="darkgoldenrod2", "rail"="cyan"), name="Mode")+
  xlab("Cost")+
theme_bw()
  
```



Create a plot showing the univariate distributions of time by car and time by bus. Discuss.


```{r}
grafica_time <- ggplot(Mode)+
  geom_density(aes(time.car, fill="car"), color="black", alpha=0.5)+
  geom_density(aes(time.carpool , fill="carpool"), color="black", alpha=0.5)+
  geom_density(aes(time.bus, fill="bus"), color="black", alpha=0.5)+
  geom_density(aes(time.rail, fill="rail"), color="black", alpha=0.5)+
  scale_fill_manual(values=c("car"= "firebrick", "carpool"="dodgerblue", "bus"="darkgoldenrod2", "rail"="cyan"), name="Mode")+
  xlab("Time")+
theme_bw()


ggplotly(grafica_time)  
# El tiempo en bus tiende a ser mayor y más disperso que en carro.
# Car tiene una distribución más concentrada, lo que sugiere mayor consistencia en tiempos.
# En general, los viajes en bus toman más tiempo que en carro.
```

How do choices relate to cost by the different modes?

```{r}
varnum <- Mode |> dplyr::select(where(is.numeric))
varnum <- names(varnum)

for (var in varnum) {
grafica_box <- ggplot(Mode)+
  geom_boxplot(aes(choice, eval(as.name(var)), fill=choice), show.legend=FALSE)+
  ylab(var)+
  theme_bw()
  
  print(grafica_box)
# Las personas tienden a evitar los modos más costosos.
# Car, aunque es el más caro, es el más elegido, lo que sugiere influencia de otros factores como comodidad o tiempo.
# Los modos más económicos como bus tienen menor preferencia, lo que refleja compensaciones entre costo y calidad del viaje.
    
}



```








